---
title: "OBI - Data (& Platform) Explorer"
author: "CivicDataLab"
date: "04/02/2020"
output: html_document
---


## Exploring the Union Budget allocation for the Ministry of Environment, Forests and Climate Change** for the year 2020-21

```{r Load Libraries, echo=FALSE, warning=TRUE}
packages_required <- c("ckanr","stringr","dplyr","DT","listviewer","ggplot2","ggalt")
lapply(packages_required, require, character.only=TRUE)
```

The objective of this notebook is to demonstrate the exploration of the [OpenBudgetsIndia](openbudgetsindia.org/) (OBI) platform right from our favourite programming environment. 

We will use the [ckanr](https://github.com/ropensci/ckanr) package developed by the amazing folks at [ROpenSci](https://ropensci.org/). This is used to read and update data from [CKAN](https://ckan.org/) instanes. Since, OBI is yet another CkAN instance, this packages works perfectly to read and explore (and write as well) data from and to the [Open Budgets India](openbudgetsindia.org/) platform. 

```{r ckan-setup}
org_url <- "https://openbudgetsindia.org/"
ckanr::ckanr_setup(url = org_url)
org_connect <- ckanr::src_ckan(url = org_url)

## Check connection
ckanr::ckan_version()
```

```{r explore-organisation}
ministry_details <- ckanr::organization_show(id = "ministry-of-environment-forests-and-climate-change",include_datasets = TRUE)

## Explore Organisation details
# listviewer::jsonedit(ministry_details)
ministry_details$description[[1]] %>% stringr::str_to_sentence() %>% cat()
```

```{r explore datasets in an org}
all_datasets <- purrr::map(ministry_details$packages,'title') %>% unlist()
```

```{r fetch package details}

# Get the package ID
data_2020_id <- ministry_details$packages[[1]]$id
pacakage_details <- package_show(data_2020_id)

# Explore package
# listviewer::jsonedit(pacakage_details)

# Get data formats
format_details <- purrr::map_df(pacakage_details$resources,`[`,c('format','id'))

print(format_details)
```

```{r get data from resource}
# Get data from CSV
res_id <- format_details$id[format_details$format == 'CSV']
# budget_data <- dplyr::tbl(src = org_connect$con, from = res_id) %>% as_tibble(.)

# data.table::fwrite(budget_data,file = "data/ministry-of-environment-forests-and-climate-change-2020.csv")
budget_data <- readr::read_csv("data/ministry-of-environment-forests-and-climate-change-2020.csv")

# Get column names
names(budget_data)
```

```{r Prepare data for analysis}
cols_required <- c('Particulars', 'Actual 2018-2019 Total', 'Revised 2019-2020 Total' ,'Budget 2020-2021 Total')
data_for_analysis <- budget_data[,cols_required]

# Clean it a bit .. 

# Remove missing values
#   - Remove rows with missing value for 'Particulars' col
#   - Convert the budget cols to numeric
#   - Convert NA to 0 in budget cols
#   - Remove rows where the sum of all budget cols is 0
# Remove whitespaces from the 'Particulars' col
# Remove the Grand Total Columns
data_for_analysis <- data_for_analysis[!is.na(data_for_analysis$Particulars),]
numeric_fields <- c('Actual 2018-2019 Total','Revised 2019-2020 Total','Budget 2020-2021 Total')
data_for_analysis[,numeric_fields] <- sapply(data_for_analysis[,numeric_fields], as.numeric)
data_for_analysis[is.na(data_for_analysis)] <- 0 
data_for_analysis  <- data_for_analysis[(data_for_analysis$`Actual 2018-2019 Total` + data_for_analysis$`Revised 2019-2020 Total` + data_for_analysis$`Budget 2020-2021 Total`) > 0,]
data_for_analysis$Particulars <- stringr::str_trim(data_for_analysis$Particulars)
data_for_analysis <- data_for_analysis[!data_for_analysis$Particulars %in% c('Grand Total'),]

# Add new column for the difference in budget from 2018-19 to 2021-21
data_for_analysis$`Budget Difference` <- data_for_analysis$`Budget 2020-2021 Total` - data_for_analysis$`Actual 2018-2019 Total`
```

```{r Table for exploration}
DT::datatable(data.table::data.table(data_for_analysis), rownames = FALSE,
    options = list(
      DT::
      searching=TRUE, 
      lengthChange=FALSE, 
      ordering=FALSE, 
      autoWidth=TRUE, 
      bPaginate=TRUE, 
      bInfo=TRUE, 
      paging=TRUE
    ))
```


Let's check how the allocation has changed from 2018-19 to 2020-21. 

```{r Analysis - Compare the actuals from 2018-19, to the planned budget for 2020-21}
top_20_gain <- data_for_analysis %>% arrange(`Budget Difference`) %>% tail(20)
bottom_20_loss <- data_for_analysis %>% arrange(`Budget Difference`) %>% head(20) 
```


```{r Graph for top 20 - Gain}
graph_for_analysis <- top_20_gain
graph_for_analysis$Particulars <- factor(graph_for_analysis$Particulars, levels = graph_for_analysis$Particulars)
top_20 <- ggplot(graph_for_analysis, aes(x=`Actual 2018-2019 Total`, xend=`Budget 2020-2021 Total`, y=Particulars, group=Particulars)) + 
        geom_dumbbell(color="#0e668b", 
                      size=0.75, 
                      point.colour.l="#0e668b") + 
        scale_x_continuous() + 
        labs(x=NULL, 
             y=NULL, 
             title="Budget Allocation (In Crores): 2018/19 vs 2020/21", 
             subtitle="Top 20 departments in terms of increase in budget allocation", 
             caption="Source: OpenBudgetsIndia") +
        theme(plot.title = element_text(hjust=0.5, face="bold"),
              plot.background=element_rect(fill="#f7f7f7"),
              panel.background=element_rect(fill="#f7f7f7"),
              panel.grid.minor=element_blank(),
              panel.grid.major.y=element_blank(),
              panel.grid.major.x=element_line(),
              axis.ticks=element_blank(),
              legend.position="top",
              panel.border=element_blank())
top_20
```

```{r Graph for bottom 20 - Loss}
graph_for_analysis <- bottom_20_loss
graph_for_analysis$Particulars <- factor(graph_for_analysis$Particulars, levels = graph_for_analysis$Particulars)
bottom_20 <- ggplot(graph_for_analysis, aes(x=`Actual 2018-2019 Total`, xend=`Budget 2020-2021 Total`, y=Particulars, group=Particulars)) + 
        geom_dumbbell(color="#0e668b", 
                      size=0.75, 
                      point.colour.l="#0e668b") + 
        scale_x_continuous() + 
        labs(x=NULL, 
             y=NULL, 
             title="Budget Allocation (In Crores): 2018/19 vs 2020/21", 
             subtitle="Bottom 20 departments in terms of increase in budget allocation", 
             caption="Source: OpenBudgetsIndia") +
        theme(plot.title = element_text(hjust=0.5, face="bold"),
              plot.background=element_rect(fill="#f7f7f7"),
              panel.background=element_rect(fill="#f7f7f7"),
              panel.grid.minor=element_blank(),
              panel.grid.major.y=element_blank(),
              panel.grid.major.x=element_line(),
              axis.ticks=element_blank(),
              legend.position="top",
              panel.border=element_blank())
bottom_20
```
